### 2014-10-02 paper.tex
added some citations

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to the high cost and low throughput
of sequencing expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into the complexity of transcriptomes (@cite).
It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.
Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

@@

and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
incomplete nature of the current chicken genome assembly, which is lacking
approximately 5\% of its true gene content (@cite).  Interestingly, over a
hundred regions were present {\em only} in the local assemblies,
suggesting that the local assemblies might be recovering additional exons.
Significant numbers of unique regions from both global and local assemblies
showed homology to the mouse genome, indicating that at least some of these
unique sequences represented real sequence content.
Figure~\ref{global_vs_local} shows an example of different isoforms detected by
the two assembly methods.

@@

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse transcription (@cite).  Thus the
merged gene models produced by Gimme represent the significant majority of the
assemblable data.

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both
methods to the ENSEMBL gene annotations, which are produced by a
pipeline that incorporates de novo gene prediction and homology-based
approaches as well as expression data (@cite).

>>>>

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to the high cost and low throughput
of sequencing expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into the complexity of transcriptomes~\cite{nagalakshmi2008transcriptional,mortazavi2008mapping,pan2008deep,Wang:2008ea}.
It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently.  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.
Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

@@

and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
incomplete nature of the current chicken genome assembly, which is lacking
approximately 5\% of its true gene content~\cite{hillier2004sequence}.
Interestingly, over a hundred regions were present {\em only} in the local
assemblies, suggesting that the local assemblies might be recovering additional
exons.  Significant numbers of unique regions from both global and local
assemblies showed homology to the mouse genome, indicating that at least some
of these unique sequences represented real sequence content.
Figure~\ref{global_vs_local} shows an example of different isoforms detected by
the two assembly methods.

@@

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse
transcription~\cite{martin2011next}.  Thus the merged gene models produced by
Gimme represent the significant majority of the assemblable data.

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both
methods to the ENSEMBL gene annotations, which are produced by a
pipeline that incorporates de novo gene prediction and homology-based
approaches as well as expression data~\cite{flicek2013ensembl}.
